---
title: "Vascular plants analysis for NI"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NI_plants_analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette documents the workflow for downloading and analysing occurrence data for 17 species of vascular plants across Norway, and subsequently using the results to calculate spatially-explicit indicator values to enter into the Nature Index (NI) database. 

Code contained within the `NIplants` package, as well as the Workflow outlined in this vignette, are adapted from scripts by Olav Skarpaas: https://github.com/skarpaas/NI2020-plants
A summary of changes that have been made relative to the source material is provided in Appendix 1 at the end of the vignette. 

For more information on the modelling framework and its assumptions and compromises, as well as a discussion of results from a biological viewpoint, refer to original documentation by Olav Skarpaas: https://github.com/skarpaas/NI2020-plants/blob/main/NI-2020-plants-report.docx


## General setup

### Installing and loading packages 

We start by loading all required packages. This includes both `NIplants` and its dependencies: 

```{r setup, message = FALSE, warning = FALSE, eval = FALSE}
library(NIplants)
library(NIcalc)
library(magrittr)
library(rgbif)
library(rgdal)
library(sp)
library(raster)
library(rio)
library(mgcv)
```

Install and/or update packages as needed. 
The up-to-date version of the `NIplants` package has to be installed from GitHub as it is not on CRAN (yet):

```{r, eval = FALSE}
devtools::install_github("DrMattG/NI_plants", ref = "InProgress_CRN")
```

### Setting paths and directories

Next, we define the paths to the analysis directory (`analysis_path`) and and the directory to store the raw data (`data_path`). Note that the paths will be different depending on whether one runs the analysis from a local computer or NINA's RStudio server.   
```{r, eval = FALSE}
# Run locally
analysis_path <- "P:/41201612_naturindeks_2021_2023_database_og_innsynslosning/Karplanter_Dataflyt/NI2025_plants"
data_path <- "P:/41201612_naturindeks_2021_2023_database_og_innsynslosning/Karplanter_Dataflyt/GBIF_Data" # Local

# Run from server
#analysis_path <- "/data/P-Prosjekter/41201612_naturindeks_2021_2023_database_og_innsynslosning/Karplanter_Dataflyt/NI2025_plants"
#data_path <- "/data/P-Prosjekter/41201612_naturindeks_2021_2023_database_og_innsynslosning/Karplanter_Dataflyt/GBIF_Data"
```
We then proceed to make sure that the analysis directory  contains two sub-folders named "Data" and "Results". 
```{r, eval = FALSE}
if(!dir.exists(paste0(analysis_path, "/Data"))){
  dir.create(paste0(analysis_path, "/Data"))
}
if(!dir.exists(paste0(analysis_path, "/Results"))){
  dir.create(paste0(analysis_path, "/Results"))
}
```

### Defining species to include in analyses
As per now, this analysis includes 17 species of vascular plants found in Norway. We define them via both their Norwegian names (used in the NI database, `indicators`) and their latin names (`species`).

```{r, eval = FALSE}
indicators <- c("Alm","Fjellvalmue","Greplyng","Issoleie","Kusymre","Myrtelg","Olavsstake","Prestekrage",
                "Purpurlyng","Sennegras","Solblom","Sveltstarr","Engmarihand","Hvitmyrak","Brunmyrak","Smalsoldogg","Dikesoldogg")

species <- c("Ulmus glabra","Papaver radicatum","Kalmia procumbens","Ranunculus glacialis","Primula vulgaris",
             "Thelypteris palustris","Moneses uniflora","Leucanthemum vulgare","Erica cinerea","Carex vesicaria",
             "Arnica montana","Carex pauciflora","Dactylorhiza incarnata","Rhynchospora alba","Rhynchospora fusca",
             "Drosera anglica","Drosera intermedia")
```


### Setting "switches" for executing different steps of the workflow
This workflow as a whole takes a long time to run, and one may not always want to (re-)run all the different steps. To enable skipping of steps we do not want to run/repeat, we define a set of logical "switches". If a certain switch is set to `FALSE`, the corresponding step of the workflow will be skipped.

```{r, eval = FALSE}

# Step 1: Downloading old NI data
#downloadOldNIData <- TRUE
downloadOldNIData <- FALSE

# Step 2: Downloading raw occurrence data from GBIF
#downloadGBIFData <- TRUE
downloadGBIFData <- FALSE

# Step 3: Building training datasets for model fitting
#buildTrainingData <- TRUE
buildTrainingData <- FALSE

# Step 4: Fitting GAMs
#fitGAM <- TRUE
fitGAM <- FALSE

# Step 5: Making GAM predictions
#predictGAM <- TRUE
predictGAM <- FALSE

# Step 6: Calculating indicator values
#calcIndicators <- TRUE
calcIndicators <- FALSE

# Step 7: Formatting indicator values
#formatIndicators <- TRUE
formatIndicators <- FALSE

# Step 8: Comparing new indicator values to old and/or upload to NI database
#processIndicators <- TRUE
processIndicators <- FALSE
```

## Retrieving old indicator data from NI database or local save 

We may want to start a new analysis by downloading and inspecting the indicator values that were previously uploaded to the NI database. 
`NI_plants::downloadData_NIdb()` allows to download the most recent indicator entries from the NI database. The function will ask for the user's NI database credentials (registered email address and a password).
If we have previously downloaded the old NI data and do not wish to repeat this step (`downloadOldNIData == FALSE`), we instead load a previously saved version of the data: 
```{r, eval = FALSE}
if(downloadOldNIData){
  
  # Execute function for downloading old indicator data
  oldIndicatorData <- downloadData_NIdb(species = species, indicators = indicators, 
                                        save = TRUE, save_path = paste0(analysis_path, "/Data"))

}else{
  
  # Load old indicator data from local save
  oldIndicatorData <- readRDS(paste0(analysis_path, '/Data/oldIndicatorData.rds'))
}
```

## Retrieving raw occurrence data from GBIF or local save 

The vascular plant indicators are based on occurrence data downloaded from GBIF. To get data from GBIF, one first needs to request a data download via a key. The function `NI_plants::getKey_GBIF()` takes care of that. It first extracts and prints all taxon keys relevant to the species listed in `species`, then asks for your GBIF login credentials, and finally proceeds to request a download. 
```{r, eval = FALSE}
# Execute function for setting up download key and requesting download
if(downloadGBIFData){
 download_key <- getKey_GBIF()
 message("Download key retrieved.")
 message("Wait until your GBIF download is ready (typically within 15 minutes, but can take up to 3 hours).")
 message("You should receive an email confirmation once the download is ready.")
}
```
Once the GBIF download is ready, we use `NI_plants::downloadGBIFData` to download the occurrence data into the directory we set previously (`data_path`). The function also unzips the raw data file (`occurrence.txt`) and moves it back where it belongs (i.e. in the data directory).
```{r, eval = FALSE}
if(downloadGBIFData){
  
  # Execute function to download, unzip, and import raw data (using generated key)
  sp <- downloadData_GBIF(key = download_key$key, path = data_path)
  
  # Move file from working directory into GBIF data folder
  fs::file_move("occurrence.txt", paste0(data_path, "/occurrence.txt"))
  
}
  
# Import raw data from previously downloaded occurrence file
GBIF_data <- import(paste0(data_path, "/occurrence.txt"), header = T, sep = "\t")

```
Note that `NI_plants::downloadGBIFData` also writes a .txt file named "GBIF_Data_Citation.txt" into the data directory. This file contains citation information for the dataset. 

## Building training data for each focal species 

Once we have the raw data ready, it needs to be filtered and re-formatted into species-specific training datasets on which we run the GAMs. The corresponding function `NI_plants::build_TrainingData` requires us to set the first and last year(`min.year` and `max.year`), as well as the time interval (`year.interval`), to consider. The time interval is set to 5 years since NI is calculated every 5 years. `NI_plants::build_TrainingData` then adds up all occurrences (presences only) contained within the time interval and assigns them to the starting year. 
```{r, results = FALSE, eval = FALSE}
# Set start, end, and year interval for time period of interest
min.year <- 1820
max.year <- 2020
year.interval <- 5

# Execute function for building training data
if(buildTrainingData){
  build_TrainingData(GBIF_data, species, min.year, max.year, year.interval, 
                     save_path = paste0(analysis_path, "/Data/"), save.unfiltered = TRUE)
}
```



## Analysing data with GAMs 

Next, we analyse each training dataset with a Generalized Additive Model (GAM). `NI_plants::run_GAM()` fits a Poisson GAM with additive effects of latitude, longitude, and date, as well as interactive effects of date with latitude and longitude, to the time-series of occurrence counts contained in the training data. An offset is used to account for imperfect detectability (see the original documentation by Olav Skarpaas for more details: https://github.com/skarpaas/NI2020-plants/blob/main/NI-2020-plants-report.docx)
```{r, eval = FALSE}
# Execute function to run GAM analyses
if(fitGAM){
  gam.results <- run_GAM(species = species, data_path = paste0(analysis_path, "/Data"), 
                         save = TRUE, save_path = paste0(analysis_path, "/Results"))
}
```


## Making spatial predictions from GAMs
The vascular plant indicators in NI are based off spatially explicit predictions from the GAMs fit with `NI_plants::run_GAM()`. These predictions are made by the function `NI_plants::predict_GAM()` for a selected number of years (specified in the argument `yearList`). As a minimum, predictions need to be made for the reference year (1900) and each year for which NI will be calculated (1990, 2000, 2010, 2014. 2019, etc.). The function also has a plotting option: if `plot.pdf = TRUE`, then a map of predicted occurrences (proxy for abundance) will be plotted for each species and year and saved as a pdf. 
Be aware that this is the most computationally intensive part of the workflow. It takes several days to run, and using a cluster or RStudio server is recommended (make sure you adjust the paths if necessary before running). 
```{r, eval = FALSE}
# Make a list of years for which to make predictions
yearList <- c(1900, 1950, 1990, 2000, 2010, 2014, 2019)

# Execute function to predict from GAMs and plot predictions (optional)
if(predictGAM){
  NIGAM_All.list <- predict_GAM(species = species, year = yearList, plot.pdf = TRUE, 
                                save = TRUE, wd_path = analysis_path)
}
```


## Calculating indicator values from model predictions
Calculation of the indicator values for vascular plants involves calibrating the abundance predictions relative to an a-priori set reference value, and subsequently scaling them to the spatial resolution appropriate for entering into the the NI database. This process can be done using `NI_plants::calculate_IndicatorValues`.
At present -- and consistent with previous iterations of NI -- predicted abundance in year 1900 is treated as the reference value. However, this will likely be changed prior to the implementation of NI2025. 
```{r, eval = FALSE}
# Execute function to calculate indicator values
if(calcIndicators){
  newIndicatorData <- calculate_IndicatorValues(species = species, year = yearList, 
                                                referenceYear = 1900,
                                                save = TRUE, wd_path = analysis_path)
}
```


## Formatting data for import to NI database 
Each individual indicator value should have a measure of uncertainty. Our indicator values here get an estimate of uncertainty from the modelling procedure (GAM fitting), but this has to be expressed in a format that is compatible with the NI database. `NI_plants::create_DistObjects()` therefore transforms our new indicator data (`newIndicatorData`) into distribution objects that can be uploaded to the NI database using the functionalities of the `NIcalc` package. 
```{r, eval = FALSE}
# Execute function to create and format distribution object data
if(formatIndicators){
  updatedIndicatorData <- create_DistObjects(species = species, save = TRUE, wd_path = analysis_path)
}
```


## Comparing updated indicator data and uploading to NI database 
The final step of the workflow is double-checking our new indicator data and subsequently uploading it to the NI database. Both of these tasks are handled by `NI_plants::uploadData_NIdb` and are initiated separately depending on the function's `mode` argument. If we set `mode = "compare"`, `NI_plants::uploadData_NIdb` connects to the NI database (requires your credentials), downloads the latest version of the indicator data stored there, and compares that data to the newly calculated indicator data on a per-species basis by calculating the ratio of the old and new indicator value (as old/new). It then returns a summary of the comparison containing means and percentiles of the relative change in indicator values.  
```{r, eval = FALSE}
# Exectue function to compare updated indicator data to previous
if(processIndicators){
  uploadData_NIdb(species = species, speciesList = species, mode = "compare", newdata_path = paste0(analysis_path, "/Data"))
}
```

Once we have compared and are satisfied with our newly calculated indicator data, we can upload it to the NI database by setting `mode = "overwrite"` in `NI_plants::uploadData_NIdb`. Note that doing this will overwrite the data currently contained in the database, and should only be done when everything has been double-checked and confirmed to be ready. After entering your credentials, you will also be prompted to confirm that you intend to overwrite data.

```{r, eval = FALSE}
# Execute function to write updated indicator data to NI database
if(processIndicators){
  uploadData_NIdb(species = species, speciesList = species, mode = "overwrite", , newdata_path = paste0(analysis_path, "/Data"))
}
```


## Appendix 1: Changes in workflow and code relative to source material

Relative to the source material provided by Olav Skarpaas (https://github.com/skarpaas/NI2020-plants), the following adaptations and changes have been made to code and workflow: 

  * All scripts have been collected and re-organised into an R package (`NI_plants`)
  * Function calls and operations in all scripts have been adapted/udated for compatibility with recent versions of R package dependencies
  * The download query to GBIF has been altered to improve efficiency. Instead of downloading occurrences for ALL plant species in Norway, the function `downloadData_GBIF` now only downloads data for the relevant species. This avoids problems with unzipping data files > 4GB and reduces memory requirements during processing. 
  * For assembling data sets for model fitting (function `build_TrainingData`), filtering is now done under consideration of the variables `species` and `acceptedTaxonKey` in additon to `taxonKey`. This avoids the dropping of records for certain subspecies and species that changed name. 
  * Datasets for fitting models are now built based on recorded presences only. Recorded absences contained in the GBIF data (`occurrenceType == "ABSENCE"`) are dropped when preparing datasets (function `build_TrainingData`). 
